---
title: "An Introduction to ActivityProfileR"
author: "Christian Pascual"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to ActivityProfileR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Purpose

The purpose of this demo is to walk through how to use the functionality of our package. This package creates the functional ANOVA test described in *Nonparametric comparisons of activity profiles from wear- able device data* by Chang et. al. Our demo will walk through how to set up your data, do some exploratory analyses and perform the test.

# Set Up

```{r, warning =  FALSE, message = FALSE }
library(tidyverse)
library(ActivityProfileR)
```

# Simulating Data

For this demo, we will generate accelerometer data for three independent groups of subjects. Accelerometer data is simulated by taking the positive components of a Ornstein-Uhlenbeck process. The `prepare_sim_data()` function will generate the data, given some group parameters. Each group's set of parameters should be given as a list with the following components:

- `n_subj`: how many subjects should be a part of this group?
- `group_id`: a number that uniquely identifies the group in the whole sample
- `n_points`: how many accelerometer data points should we generate for each person?
- `scaling`: what factor should we scale the resulting accelerometer data?
- `thetas`: a 3-length vector describing what thetas should be passed into `ESGtoolkit::simdiff`, which is what generates the accelerometer data

```{r}
# Set up the group parameters as a list of lists
params_by_group = list(
  list(n_subj = 70, n_points = 1000, 
       scaling = 300, thetas = c(-0.8, 2.0, 2.31)),
  list(n_subj = 100, n_points = 1000, 
       scaling = 300, thetas = c(-0.402, 1.0, 1.65)),
  list(n_subj = 130, n_points = 1000, 
       scaling = 300, thetas = c(-0.201, 0.5, 1.18))
)

data = params_by_group %>% 
  prepare_sim_data
```

`prepare_sim_data()` creates a simulated dataset of activity for a prescribed number of groups. Activity counts are simulated through an Ornstein-Uhlenbeck process. We can take a look at an example of a subject's simulated activity data below:

```{r}
data %>% 
  unnest(Xt) %>% 
  filter(subject_id == 1, group == 1) %>% 
  mutate( t = 1:1000 ) %>% # same as n_points in group_params above
  ggplot(aes(x = t, y = Xt)) +
  geom_line() +
  labs(
    title = "Sample of a subject's simulated activity data",
    x = "Time (t)",
    y = "Activity (steps)"
  )
```

`prepare_sim_data()` uses the index in the `params_by_group` list to designate a group ID column. `prepare_sim_data()` uses `group` to name the group ID column. Activity data is put in the `Xt` column as a nested tibble. `ActivityProfileR` assumes that any **user-specified data** has the same form as the  `data` created by `prepare_sim_data()`.

# Producing The Activity Profle $T(\alpha)$

The functional ANOVA test described in the original manuscript takes the original activity data, denoted as $X(t)$, and converts them into activity profiles, denoted $T(\alpha)$. Activity profiles are monotonically decreasing functions, where a single point represents the proportion of data points greater than a given activity value.

```{r}
processed_data = data %>% 
  add_helper_columns(group_col = group) %>% 
  add_activity_profile(group_col = group, 
                       activity_col = Xt,
                       grid_length = 100, 
                       quantiles = c(0.05, 0.95))
```

The `add_helper_columns()` function adds some columns that help in the calculation of the statistical test. Specifically, it adds a column `n` that represents how many subjects have a specific group ID and column `gamma`, which represents the proportion that a group ID is of the entire sample.

The `add_activity_profile()` constructs the activity profiles for each subject. These activity profiles depend on the activity values across all groups, so this information needs to be supplied to the function. Thanks to tidyeval, you do not need to specify the column names as a string. In the simulated dataset, the group column is represented as `group` and activity column as `Xt`, so this should be adjusted to the input data as needed. `add_activity_profile()` also adds a column `ai`, which stands for activity index. Since activity values can differ across dataset, these activity indices are used as a placeholder. `grid_length` and `quantiles` are optional arguments that you can supply to `add_activity_profile` if you wish to change the amount of activity values to use or change the quantiles kept. The first value, `ai = 1`, will represent the 5th percentile of all activity values in the entire sample, while the last `ai = 100` will represent the 95th percentile (per the specifications of the manuscript).

The end result is that activity profiles have been created for each subject across all groups:

```{r, echo = FALSE}
head(processed_data)
```

# Visualizing $T(\alpha)$

We can visualize the activity profiles for each group by unnesting the data and plotting by group.

```{r}
processed_data %>% 
  unnest(c(Ta, ai)) %>% 
  ggplot(aes(x = ai, y = Ta, group = subject_id, color = factor(group))) + 
  geom_line(alpha = 0.3) + 
  facet_grid(group ~ .) +
  theme(legend.position = "bottom") +
  labs(
    title = "Activity profiles for each subject by group",
    x = "Activity index alpha",
    y = "Proportion of data that exceeds\nactivity at index alpha"
  )
```

# Calculating Functional ANOVA Test

The functions above are used internally in the `EL_test()`, which calculates the statistical test. 

```{r, message = FALSE, warning = FALSE }
start = Sys.time()
el = processed_data %>% 
  EL_test(id_col = subject_id,
          group_col = group, 
          activity_col = Xt)
end = Sys.time()
paste0("Regular version of EL_test() took ", 
       round(end - start, 1), 
       " minutes to run.") %>% print
```

The output of `EL_test()` is a list of values associated with the functional ANOVA test. 

- `sup_test`: This is the supremum of all of the $-2log(\alpha)$ values calculated for each activity index $\alpha$.
- `sup_EL_crit`: This is the 95% quantile of the bootstrap statistics calculated on the data.
- `out_sup_pval`: This represents the proportion of bootstrap test statistics were greater than `sup_test`.
- `neg2logRa_tests`: These are the $-2log(\alpha)$ values calculated for each activity index $\alpha$.
- `bootstrap`: These are all of the bootstrap datasets created from the original data.

```{r}
# Investigate the output of the test
el$sup_test
el$sup_EL_crit
el$out_sup_pval
```
